---
title: "MSMB-Chapter4-Mixture Models"
author: "Aleeza Gerstein"
date: '2019-10-20'
output:
  rmarkdown::pdf_document:
  fig_caption: yes
includes:
  in_header: preamble-latex.tex
header-includes:
  - \usepackage{color}
  - \usepackage{mathtools}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("mixtools"))
```

# Mixture Models

## Finite mixtures

```{r twocoins, fig.keep = 'high', fig.cap = "Histogram of 10,000 random draws from a fair mixture of two normals. The left hand part of the histogram is dominated by numbers generated from (A), on the right from (B).", fig.width = 3.5, fig.height = 3.5}
coinflips = (runif(10000) > 0.5)
table(coinflips)
oneFlip = function(fl, mean1 = 1, mean2 = 3, sd1 = 0.5, sd2 = 0.5) {
  if (fl) {
    rnorm(1, mean1, sd1)
  } else {
    rnorm(1, mean2, sd2)
  }
}
fairmix = vapply(coinflips, oneFlip, numeric(1))
ggplot(tibble(value = fairmix), aes(x = value)) +
  geom_histogram(fill = "purple", binwidth = 0.1)
```

---

\textcolor{red}{Question 4.1 How can you use R's vectorized syntax to remove the `vapply` loop and generate the `fairmix` vector more efficiently?}
\textcolor{red}{Question 4.2 Using your improved code, perform one million flips and make a histogram with 500 bins}

```{r}
sd <- 0.25
coinflips_mil = (runif(1000000) > 0.5)
mean_mil <- ifelse(coinflips_mil, 1, 3)
values <- rnorm(length(coinflips_mil), mean_mil, 0.25)
ggplot(tibble(value = values), aes(x = value)) +
  geom_histogram(fill = "purple", bins = 500)
```

---

---

\textcolor{red}{Question 4.3 Plot a histogram only of those values for which coinflips is `TRUE`. Hint: use `y = ..density ..` in the call to `aes` to specify that the vertical axis shows the proportion of counts and set the binwidth to 0.01. (b) Overlay the line corresponding to $\phi(z)$}

```{r}
fair <- tibble(coinflips_mil, values)
ggplot(filter(fair, coinflips_mil), aes(values)) +
  geom_histogram(aes(y = ..density..), fill = "purple", binwidth = 0.01) +
  stat_function(fun = dnorm, args = list(mean = 1, sd = 0.25), color="red")
```

---

You can write the formula for the density of all values as a sum of the two densities.

```{r}
means = c(1, 3)
sds   = c(0.25, 0.25)

fairtheory = tibble(
  x = seq(-1, 5, length.out = 1000),
  f = 0.5 * dnorm(x, mean = means[1], sd = sds[1]) +
    0.5 * dnorm(x, mean = means[2], sd = sds[2]))
ggplot(fairtheory, aes(x = x, y = f)) +
  geom_line(color = "red", size = 1.5) + ylab("mixture density")
```

---

\textcolor{red}{Question 4.4 Figure 4.4 is a histogram of a fair mixture model of two normals with the same variances. Can you guess the two mean parameters of the component distribution? Hint: You can use trial and error, and simulate variance mixtures to see if you can make a matching histogram.}


```{r}
means <- c(1, 2)
coinflips_mil = (runif(10000) > 0.5)
tib <- tibble(mean_mil = ifelse(coinflips_mil, means[1], means[2]))
tib$values <- rnorm(length(coinflips_mil), tib$mean_mil, sqrt(.5))
#br = with(mystery, seq(min(values), max(values), length.out=30))
br = with(tib, seq(min(values), max(values), length.out=30))

ggplot(tib, aes(x = values)) +
  geom_histogram(fill = "purple", binwidth = 0.02, breaks=br) 
```

\textcolor{blue}{Tell us much earlier why you are setting it up this way with means <- c(1, 2) etc.}

---

\textcolor{red}{Question 4.5 Suppose we have two unfair coins with probability of heads $p_1$ = 0.125 and $p_2$ = 0.25. With probability $\pi$ we pick coin 1 and with probability 1-$\pi$ we pick coin 2. Toss that coin twice and record the number of heads K. 
(a) Simulate 100 instances of this procedure with $\pi = \frac{1}{8}$ and compute the contingency table of K.
(b) Do the same with $\pi = \frac{1}{4}$}
(c) If you were not told $p_1$, $p_2$ and $\pi$ could you infer them from the contingency table?

```{r}
ps <- c(0.125, 0.25)
choose <- function(pi){ 
  coin <- (runif(1) > pi)
  if(coin) t <- rbinom(2, 1, ps[1])
  else t <- rbinom(2, 1, ps[2])
  t
}

K <- c()
for(i in 1:100){
  t <- choose(1/8)
  K <- append(K, sum(t))
}
table(K)

K2 <- c()
for(i in 1:100){
  t <- choose(1/4)
  K2 <- append(K2, sum(t))
}
table(K2)
```

There is an easier way. 

```{r}
probHead = c(0.125, 0.25)
for (pi in c(1/8, 1/4)) {
  whCoin = sample(2, 100, replace = TRUE, prob = c(pi, 1-pi))
  K = rbinom(length(whCoin), size = 2, prob = probHead[whCoin])
  print(table(K))
}
```

---

```{r}
y = c(rnorm(100, mean = -0.2, sd = 0.5),
      rnorm( 50, mean =  0.5, sd =   1))
gm = normalmixEM(y, k = 2, lambda = c(0.5, 0.5),
                 mu = c(-0.01, 0.01), sigma = c(1, 1))
gm$lambda
gm$mu
gm$sigma
gm$loglik
```

```{r}
masses = c(A =  331, C =  307, G =  347, T =  322)
probs  = c(A = 0.12, C = 0.38, G = 0.36, T = 0.14)
N  = 7000
sd = 3
nuclt   = sample(length(probs), N, replace = TRUE, prob = probs)
quadwts = rnorm(length(nuclt),
                mean = masses[nuclt],
                sd   = sd)
ggplot(tibble(quadwts = quadwts), aes(x = quadwts)) +
  geom_histogram(bins = 100, fill = "purple")
```

---

\textcolor{blue}{Question 4.9: Repeat the simulation with `N=1000` nucleotide measurements. What do you notice in the histogram?}

```{r}
masses = c(A =  331, C =  307, G =  347, T =  322)
probs  = c(A = 0.12, C = 0.38, G = 0.36, T = 0.14)
N  = 1000
sd = 3
nuclt   = sample(length(probs), N, replace = TRUE, prob = probs)
quadwts = rnorm(length(nuclt),
                mean = masses[nuclt],
                sd   = sd)
ggplot(tibble(quadwts = quadwts), aes(x = quadwts)) +
  geom_histogram(bins = 100, fill = "purple")
```

\textcolor{red}{Question 4.10 What happens when N = 7000 but the standard deviation is 10?}

```{r}
masses = c(A =  331, C =  307, G =  347, T =  322)
probs  = c(A = 0.12, C = 0.38, G = 0.36, T = 0.14)
N  = 7000
sd = 10
nuclt   = sample(length(probs), N, replace = TRUE, prob = probs)
quadwts = rnorm(length(nuclt),
                mean = masses[nuclt],
                sd   = sd)
ggplot(tibble(quadwts = quadwts), aes(x = quadwts)) +
  geom_histogram(bins = 100, fill = "purple")
```

\textcolor{red}{Plot he theoretical density curve for the distribution simulated in Figure 4.9}

```{r}
masses = c(A =  331, C =  307, G =  347, T =  322)
probs  = c(A = 0.12, C = 0.38, G = 0.36, T = 0.14)
N  = 7000
sd = 3
nuclt   = sample(length(probs), N, replace = TRUE, prob = probs)
quadwts = rnorm(length(nuclt),
                mean = masses[nuclt],
                sd   = sd)

fairtheory <- tibble(
  x = seq(295, 360, length.out = 1000),
  f = 0.12*dnorm(x, mean = 331, sd = 3) +
      0.38*dnorm(x, mean = 307, sd = 3) +
      0.36*dnorm(x, mean = 347, sd = 3) +
      0.14*dnorm(x, mean = 322, sd = 3) 
)
ggplot(fairtheory, aes(x = x, y = f)) +
  geom_line(color ="red", size = 1.5) +
  ylab("mixture density")
```
